Classless Java

### 2015-11-17 paper/sections/Overview.tex

\section{Overview}\label{sec:ep}

@@

\subsection{A Running Example: Animals}
\marco{fig1 is good but is very big, not sure if it worth its space. If we keep it, we need to do it with a vectorial graphic, or better inside latex (for example with tikz)}
To propose a standard example, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ as a kind of \Q@Animal@ with the skills of both \Q@Horse@s and \Q@Bird@s.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argue against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

%Suppose we want to model an animal system as shown in Fig~\ref{fig:pegasus}. In
%order to model \texttt{Pegasus}, \emph{multiple inheritance} is clearly needed
%in this example. Therefore, in Java we use interfaces instead of classes. An
%\texttt{Animal} has a \texttt{point} field, indicating the current position of
%the animal. \texttt{Horse} and \texttt{Bird} are subtypes of \texttt{Animal},
%with methods \texttt{run()} and \texttt{fly()}, respectively. Note that
%\texttt{Bird} can fly, therefore refines the type of \texttt{point} from
%\texttt{Point2D} to \texttt{Point3D}. Pegasus (one of the best known creatures
%in Greek mythology) can not only \emph{run} but also \emph{fly}! This is the
%place where \emph{``multiple inheritance''} is needed, because \texttt{Pegasus}
%need to obtain \texttt{fly} and \texttt{run} functionality from both
%\texttt{Horse} and \texttt{Bird}.
%---------------
% We can put more in dept discussion while discussing the classes in details.
%--------------
\subsubsection{\Q@Point2D@: simple immutable data with fields}
Since in ClassLess Java there are no classes, our \Q@Point2D@ must be an interface, as for example:
\begin{lstlisting}
interface Point2D{ int x(); int y();}
\end{lstlisting}
That is, since Java disallows fields inside interfaces,
 we simulate state by abstract methods.
By using the new Java8 \Q@default@ methods, we can easily add implemented methods inside the interface. For example:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  default float distanceFromOrigin(){ return Math.sqrt(x()*x()+y()*y());}
  }
\end{lstlisting}
resemble a class that have fields (getters) and methods.
However, how can we create an instance of \Q@Point2D@ without using classes? 
we can just provide an implementation for \Q@x()@ and \Q@y()@ as in
\begin{lstlisting}
new Point2D(){ public int x(){return 4;} public int y(){return 2;}}
\end{lstlisting}
However we are not required to use this cumbersome syntax\footnote{Available in Java from version ...} for every object allocation.
As programmes do, for ease or reuse, we can encapsulate boring long repetitive code in a method. A static factory method in this case is appropriate:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
  }
\end{lstlisting}
This obvious constructor code can be automatically generated by our \mixin annotation.

By annotating the interface \Q@Point2D@, the annotation will generate the shown static method \texttt{of}, mimicking the functionality of a simple minded constructor:
By looking the the methods that need implementation it first detects what are the fields, then 
generate an \Q@of@ method with one argument for each of them.

@@

More precisely, a field or factory parameter is generated for every no-args method requiring implementation whose name does not have special meaning\footnote{Formal definition of \mixin behaviour is provided later.}.
An example of code using \Q@Point2D@ is 
\Q@Point2D.of(42,myPoint.y())@
where we return a new point, using \Q@42@ as x-coordinate, and 
taking all the other informations (only \Q@y@ in this case) from another point.
This pattern is very common when programming with immutable data-structures; it is so common that we decided to support it in our code generation as \Q@with-@ methods, that is:
\begin{lstlisting}
@Mixin interface Point2D{ int x();    int y();
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
is expanded by our annotation into

@@

\subsubsection{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
\Q@Points2D@ are mathematical entities, thus we chosen to use immutable data structure to model them; however animals are real world entities, and when an animal moves, it is the same animal that have now a different location. We model this with mutable state.
%Now we proceed to define \texttt{Animal} with \texttt{point} ``member
%field''.
%%Not again, before we used only getters!
% Again, we model this member field with getter and setter methods:
\lstinputlisting[linerange=64-67]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL
\marco{TODO: change point into ``location''}
Here we declare abstract getter and setter for the mutable ``field'' \Q@location@.
As you see, we do not apply the \mixin annotation. This is morally equivalent to an abstract class in full Java; that is we do not provide a convenient way to instantiate it.

@@

\subsubsection{\Q@Bird@:  field type refinement}
\Q@Bird@ needs a 3D location, thus first of all we define \Q@Point3D@ and
as follows:
\lstinputlisting[linerange=56-60]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D

@@

@@

Many here would think that a language extension is needed.
Instead, with our language \emph{restriction} the problem can be easily challenged\footnote{Often in programming languages ``freedom is slavery''.}:

@@

Note how
\begin{itemize}
\item by covariant method overriding we refine the type of our field
\item by \emph{overloading} we define a new setter with the more precise type
\item by default method we provide an implementation for the setter called with the old signature.
\end{itemize}

@@

\subsubsection{\Q@Pegasus@ and multiple Inheritance}
The \emph{``multiple inheritance''} case appears at interface
\texttt{Pegasus}.
Pegasus (one of the best known creatures
in Greek mythology) can not only \emph{run} but also \emph{fly}! 
It would be handy if they could obtain \texttt{fly} and \texttt{run} functionality from both \texttt{Horse} and \texttt{Bird}.

>>>>

\section{A Running Example: Animals}\label{sec:ep}

@@

\marco{fig1 is now better. Where we cite it in the text? can we make it so that is more "linear"? point2d <- point3D,   animal <- (horse over bird ) <- pegasus}
To propose a standard example, we show \Q@Animal@s with a two dimensional \Q@Point2D@ representing their \Q@location@.
Some kinds of animals are \Q@Horse@s and \Q@Bird@s.
Birds can \Q@fly@, thus their location need to be a three dimensional \Q@Point3D@.
Finally, we model \Q@Pegasus@ as a kind of \Q@Animal@ with the skills of both \Q@Horse@s and \Q@Bird@s.\footnote{
Some research argues in favour of using subtyping for modelling taxonomies, other research argue against this practice, we do not wish to take sides in this argument, but to provide an engaging example.}

%Suppose we want to model an animal system as shown in Fig~\ref{fig:pegasus}. In
%order to model \texttt{Pegasus}, \emph{multiple inheritance} is clearly needed
%in this example. Therefore, in Java we use interfaces instead of classes. An
%\texttt{Animal} has a \texttt{point} field, indicating the current position of
%the animal. \texttt{Horse} and \texttt{Bird} are subtypes of \texttt{Animal},
%with methods \texttt{run()} and \texttt{fly()}, respectively. Note that
%\texttt{Bird} can fly, therefore refines the type of \texttt{point} from
%\texttt{Point2D} to \texttt{Point3D}. Pegasus (one of the best known creatures
%in Greek mythology) can not only \emph{run} but also \emph{fly}! This is the
%place where \emph{``multiple inheritance''} is needed, because \texttt{Pegasus}
%need to obtain \texttt{fly} and \texttt{run} functionality from both
%\texttt{Horse} and \texttt{Bird}.
%---------------
% We can put more in dept discussion while discussing the classes in details.
%--------------
\subsection{\Q@Point2D@: simple immutable data with fields}
Since in ClassLess Java there are no classes, our \Q@Point2D@ must be an interface, as for example:
\begin{lstlisting}
interface Point2D{ int x(); int y();}
\end{lstlisting}
That is, since Java disallows fields inside interfaces,
 we simulate state by abstract methods.
By using the new Java8 \Q@default@ methods, we can easily add implemented methods inside the interface. For example:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  default float distanceFromOrigin(){ return Math.sqrt(x()*x()+y()*y());}
  }
\end{lstlisting}
resemble a class that have fields (getters) and methods.
However, how can we create an instance of \Q@Point2D@ without using classes? 
we can just provide an implementation for \Q@x()@ and \Q@y()@ as in
\begin{lstlisting}
new Point2D(){ public int x(){return 4;} public int y(){return 2;}}
\end{lstlisting}
However we are not required to use this cumbersome syntax\footnote{Available in Java from version ...} for every object allocation.
As programmes do, for ease or reuse, we can encapsulate boring long repetitive code in a method. A static factory method in this case is appropriate:
\begin{lstlisting}
interface Point2D{ int x(); int y();
  static Point2D of(int x, int y){return new Point2D(){
    public int x(){return x;} public int y(){return y;}};}
  }
\end{lstlisting}
This obvious constructor code can be automatically generated by our \mixin annotation.

By annotating the interface \Q@Point2D@, the annotation will generate a variation of the shown static method \texttt{of}, mimicking the functionality of a simple minded constructor:
By looking the the methods that need implementation it first detects what are the fields, then 
generate an \Q@of@ method with one argument for each of them.

@@

More precisely, a field or factory parameter is generated for every no-args method requiring implementation whose name does not have special meaning\footnote{Formal definition of \mixin behaviour is provided later.}.
An example of code using \Q@Point2D@ is 
\Q@Point2D.of(42,myPoint.y())@
where we return a new point, using \Q@42@ as x-coordinate, and 
taking all the other informations (only \Q@y@ in this case) from another point.
This pattern is very common when programming with immutable data-structures; it is so common that we decided to support it in our code generation as \Q@with-@ methods, that is:
\begin{lstlisting}
@Mixin interface Point2D{ int x();    int y();
  Point2D withX(int val);    Point2D withY(int val);}
\end{lstlisting}
is expanded by our annotation into\footnote{
Note how we actually generate a real field \Q@int x=_x;@.
This provide a more uniform translation that can work also for mutable datastructures, where setters are required.
}

@@

\subsection{\Q@Animal@ and \Q@Horse@: simple mutable data with fields}
\Q@Points2D@ are mathematical entities, thus we chosen to use immutable data structure to model them; however animals are real world entities, and when an animal moves, it is the same animal that have now a different location. We model this with mutable state.
%Now we proceed to define \texttt{Animal} with \texttt{point} ``member
%field''.
%%Not again, before we used only getters!
% Again, we model this member field with getter and setter methods:
\lstinputlisting[linerange=64-67]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=ANIMAL
\marco{TODO: change point into ``location''}
Here we declare abstract getter and setter for the mutable ``field'' \Q@location@.
As you see, we do not apply the \mixin annotation. This is morally equivalent to an abstract class in full Java; that is we do not provide a convenient way to instantiate it.

@@

\subsection{\Q@Bird@:  field type refinement}

\subsubsection*{\Q@Point3D@ and properties updater}
\Q@Bird@ needs a 3D location, thus first of all we define \Q@Point3D@ and
as follows:
\lstinputlisting[linerange=56-60]{../UseMixinLombok/src/test/TestAnimal.java}% APPLY:linerange=POINT3D

@@

\subsubsection*{Unsatisfactory class based solutions to field type refinement}

@@

Many readers would now think that a language extension is needed.
Instead, with our language \emph{restriction} the problem can be easily challenged\footnote{Often in programming languages ``freedom is slavery''.}.
\subsubsection*{Our approach to field type refinement}

@@

In our proposed solution,
\begin{itemize}
\item by covariant method overriding we refine the field type
\item by \emph{overloading} we define a new setter with the more precise type
\item by default method we provide an implementation for the setter called with the old signature.
\end{itemize}

@@

\subsection{\Q@Pegasus@ and multiple Inheritance}
The \emph{``multiple inheritance''} case appears at interface
\texttt{Pegasus}.
Pegasus (one of the best known creatures
in Greek mythology) can not only \emph{run} but also \emph{fly}! 
It would be handy if they could obtain \texttt{fly} and \texttt{run} functionality from both \texttt{Horse} and \texttt{Bird}.

### 2015-11-17 paper/sections/RelatedWork.tex

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance.  However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is huge limitation in class based object oriented
programming, and is exasperated by interface-only programming as we propose: no
type is ever exact since classes are not part of the language.  A recent
article~\cite{Saito2013933} lights up this topic, since they propose two
features: exact statements \yanlin{do we support exact statements?} and
non-inheritable methods; both related to our work: any method generated inside
of the \Q@of@ method is indeed non-inheritable, since there is no class name to
extends from, and exact statements (a form of wild-card capture on the exact
run-time type) could capture the ``exact type'' of an object even in a
class-less environment.

@@

\subsection{Type-Directed Translations/Syntactic Sugar}
\marco{I'm tring to merge this and the next one}
Language extensions are often implemented as syntactic sugar of the base
language. For example, Scala compiler supports XML syntax in normal Scala code
directly (after Scala ?, users need to import \texttt{scala-xml} library
manually). However, this approach is hard in terms of implementation, because it
requires extending the compiler. Also, this approach does not support combining
multiple extensions into one.

@@

We can model certain types of language extensions with annotations 
only, but those extensions do not introduce new syntax: they 
merely do automatic code generation.

\subsection{Meta-programming competes with Language extensions}
The most obvious solution to add features to a language, is language extension.
They are often implemented as syntactic sugar of the base
language. For example, the Scala compiler was extended to directly supports XML syntax. However, this approach does not support combining multiple extensions into one. We are de facto creating a fork in the language, and rarely the new fork gain enough traction to become the main language release.
On this topic we mention SugarJ~\cite{erdweg11sugarj}; a Java-based extensible language allowing programmers to extend it with custom features by definitions in meta-DSLs (SDF, Stratego, etc).

On the other side, when the starting language have a flexible enough syntax and a fast and powerful enough reflection, we may just need to play with operator overloading and other language tricks to discover that the language feature we need can be expressed a a simple library in our language. For example, consider SQL alchemy in python.

@@

Here (compile/load time) code generation come at the rescue: 
if for a certain feature (\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature
(annotating a class and provide getters), then we can insert in the compilation process a tool that exam and enrich the code before compilation. No need to modify the original, source, we can work on temporary enriched source.
Java is a particular good candidate for this kind of manipulation since it already provide ways to define and integrate such tools in its own compilation process.

@@

Annotations/Code Generation/Lombok

@@

As discuss before, the user can trivially define its own \Q@of@ method, and initialize a portion of the state with default values.
However, the initialization code would not be reused/reusable, and subinterfaces would have to repeat such initialization code.
If a field has no setters, a simple alternative is to just define the ``field'' as a default method as in 
\begin{lstlisting}
@Mixin interface Box{ default int val(){return 0;} }
\end{lstlisting}
if setters are required, a possible extension of our code expansion could recognize a field if the getter is provided and the setter is required, and could generate the following code:
\begin{lstlisting}
interface Box{ 
  default int val(){return 0;} 
  void val(int _val);
  static Box of(){return new Box(){
    int val=Box.super.val();
    int val(){return val;}
    void val(int _val){val=_val;}
    };}}
\end{lstlisting}
We are unsure of the value of this solution: is very tricky, the user define a method that (contrary to our usual expectation) is actually overridden in a way that the behaviour change, but change only after the first setter is called, plus this code would cache the result instead of re-computing it every time. This can be very relevant and tricky in a non functional setting.

\subsection{Class invariants in ClassLess Java}
Since the objects are created by automatically generate methods,
another limitation of our current approach is that there is no place where the user can dynamically check for class invariants.
In Java often we see code like
\begin{lstlisting}
class Point{ int x; int y;
  Point(int x; int y){this.x=x;this.y=y; assert this.checkInvariant();}
  private boolean checkInvariant(){... x>0,y>0...}
\end{lstlisting}

@@

\subsection{Clone, toString, equals and hashCode}
Methods originally defined in \Q@Object@, as \Q@clone@ and \Q@toString@ can be supported by our approach, but they need special care. If an interface annotated with \mixin ask an implementation for \Q@clone@, \Q@toString@ \Q@equls@ or \Q@hashCode@ we can easily generate one from the fields.\footnote{In particular, for clone we can do automatic return type refinement as we do for \Q@with-@ and fluent setters. Note how this would solve most of the Java ugliness related to \Q@clone@ methods.}

>>>>

One famous attempt in this direction is provided by
\emph{MyType}~\cite{bruce1994paradigmatic}, representing the type of
\textbf{this}, changing its meaning along with inheritance.  However when
invoking a method with MyType in a parameter position, the exact type of the
receiver must be known.  This is huge limitation in class based object oriented
programming, and is exasperated by interface-only programming as we propose: no
type is ever exact since classes are not part of the language.  A recent
article~\cite{Saito2013933} lights up this topic, since they propose two
features: exact statements \yanlin{do we support exact statements?}\marco{no we do not. When is that it may seams we do?} and
non-inheritable methods; both related to our work: any method generated inside
of the \Q@of@ method is indeed non-inheritable, since there is no class name to
extends from, and exact statements (a form of wild-card capture on the exact
run-time type) could capture the ``exact type'' of an object even in a
class-less environment.

@@

\begin{comment}
\subsection{Type-Directed Translations/Syntactic Sugar}
\marco{I'm tring to merge this and the next one}
Language extensions are often implemented as syntactic sugar of the base
language. For example, Scala compiler supports XML syntax in normal Scala code
directly (after Scala ?, users need to import \texttt{scala-xml} library
manually). However, this approach is hard in terms of implementation, because it
requires extending the compiler. Also, this approach does not support combining
multiple extensions into one.

@@

We can model certain types of language extensions with annotations 
only, but those extensions do not introduce new syntax: they 
merely do automatic code generation. 
\end{comment}
\subsection{Meta-programming competes with Language extensions}
The most obvious solution to add features to a language is language extension.
They are often implemented as syntactic extensions that can be desugared to the base
language. For example, the Scala compiler was extended to directly supports XML syntax. However, this approach does not support combining multiple extensions into one. We are de facto creating a fork in the language, and rarely the new fork gain enough traction to become the main language release.
On this topic we mention SugarJ~\cite{erdweg11sugarj}; a Java-based extensible language allowing programmers to extend it with custom features by definitions in meta-DSLs (SDF, Stratego, etc).

On the other side, when the starting language have a flexible enough syntax and a fast and powerful enough reflection, we may just need to play with operator overloading and other language tricks to discover that the language feature we need can be expressed as a simple library in our language. For example, consider SQL alchemy in python.

@@

Here (compile/load time) code generation come at the rescue: 
if for a certain feature (\mixin in our case) it is possible to use the original language syntax to
\emph{express-describe} any specific instantiation of such feature
(annotating a class and provide getters), then we can insert in the compilation process a tool that exam and enrich the code before compilation. No need to modify the original source; for example we can work on temporary files.
Java is a particular good candidate for this kind of manipulation since it already provide ways to define and integrate such tools in its own compilation process: in this way there is no need of temporary files, and there is a well defined way of putting multiple extensions together.

@@

\marco{text under still to revise. Sould all the lombock discussion go in the implementation section? -- just to keep stuff together?}
Annotations/Code Generation/Lombok

@@

As discuss before, the user can trivially define its own \Q@of@ method, and initialize a portion of the state with default values.
However, the initialization code would not be reused/reusable, and subinterfaces would have to repeat such initialization code.
If a field has no setters, a simple alternative is to just define the ``field'' as a default method as in 
\begin{lstlisting}
@Mixin interface Box{ default int val(){return 0;} }
\end{lstlisting}
if setters are required, a possible extension of our code expansion could recognize a field if the getter is provided and the setter is required, and could generate the following code:
\begin{lstlisting}
interface Box{ 
  default int val(){return 0;} //provided
  void val(int _val);//provided
  static Box of(){return new Box(){//generated
    int val=Box.super.val();
    int val(){return val;}
    void val(int _val){val=_val;}
    };}}
\end{lstlisting}
We are unsure of the value of this solution: is very tricky, the user define a method that (contrary to our usual expectation) is actually overridden in a way that the behaviour change, but change only after the first setter is called, plus this code would cache the result instead of re-computing it every time. This can be very relevant and tricky in a non functional setting.

\subsection{Class invariants in ClassLess Java}
Since the objects are created by automatically generated methods,
another limitation of our current approach is that there is no place where the user can dynamically check for class invariants.
In Java often we see code like
\begin{lstlisting}
class Point{ int x; int y;
  Point(int x; int y){this.x=x;this.y=y; assert this.checkInvariant();}
  private boolean checkInvariant(){... x>0,y>0...}
\end{lstlisting}

@@

\subsection{Clone, toString, equals and hashCode}
Methods originally defined in \Q@Object@, as \Q@clone@ and \Q@toString@ can be supported by our approach, but they need special care. If an interface annotated with \mixin ask an implementation for \Q@clone@, \Q@toString@, \Q@equals@ or \Q@hashCode@ we can easily generate one from the fields.\footnote{In particular, for clone we can do automatic return type refinement as we do for \Q@with-@ and fluent setters. Note how this would solve most of the Java ugliness related to \Q@clone@ methods.}
